ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-17 notebooks/circles.qmd
appendix

>>>>

```{julia}
params = DataFrame(
    Dict(
        :n_obs => Int.(round(n_obs/10)*10),
        :epochs => epochs,
        :batch_size => batch_size,
        :n_hidden => n_hidden,
        :n_layers => length(builder.hidden),
        :activation => string(activation),
        :n_ens => n_ens,
        :lambda => string(α[3]),
        :jem_sampling_steps => jem.sampling_steps,
        :sgld_batch_size => sampler.batch_size,
        :dataname => "Circles",
    )
)
CSV.write(joinpath(params_path, "circles.csv"), params)
```

### 2023-05-17 notebooks/linearly_separable.qmd
appendix

>>>>

```{julia}
params = DataFrame(
    Dict(
        :n_obs => Int.(round(n_obs/10)*10),
        :epochs => epochs,
        :batch_size => batch_size,
        :n_hidden => n_hidden,
        :n_layers => length(builder.hidden),
        :activation => string(activation),
        :n_ens => n_ens,
        :lambda => string(α[3]),
        :jem_sampling_steps => jem.sampling_steps,
        :sgld_batch_size => sampler.batch_size,
        :dataname => "Linearly Separable",
    )
)
CSV.write(joinpath(params_path, "linearly_separable.csv"), params)
```

### 2023-05-17 notebooks/mnist.qmd
appendix

>>>>

```{julia}
params = DataFrame(
    Dict(
        :n_obs => Int.(round(n_obs/10)*10),
        :epochs => epochs,
        :batch_size => batch_size,
        :n_hidden => n_hidden,
        :n_layers => length(model_dict["MLP"].fitresult[1][1])-1,
        :activation => string(activation),
        :n_ens => n_ens,
        :lambda => string(α[3]),
        :jem_sampling_steps => jem.sampling_steps,
        :sgld_batch_size => sampler.batch_size,
        :dataname => "MNIST",
    )
)
CSV.write(joinpath(params_path, "mnist.csv"), params)
```

### 2023-05-17 notebooks/moons.qmd
rerun
appendix

@@

```{julia}
λ₁ = 0.1
λ₂ = 0.5
λ₃ = 0.5
Λ = [λ₁, λ₂, λ₃]

opt = Flux.Optimise.Descent(0.01)
use_class_loss = true

>>>>

```{julia}
params = DataFrame(
    Dict(
        :n_obs => Int.(round(n_obs/10)*10),
        :epochs => epochs,
        :batch_size => batch_size,
        :n_hidden => n_hidden,
        :n_layers => length(builder.hidden),
        :activation => string(activation),
        :n_ens => n_ens,
        :lambda => string(α[3]),
        :jem_sampling_steps => jem.sampling_steps,
        :sgld_batch_size => sampler.batch_size,
        :dataname => "Moons",
    )
)
CSV.write(joinpath(params_path, "moons.csv"), params)
```

@@

```{julia}
λ₁ = 0.25
λ₂ = 0.75
λ₃ = 0.75
Λ = [λ₁, λ₂, λ₃]

opt = Flux.Optimise.Descent(0.05)
use_class_loss = false

### 2023-05-17 notebooks/poc.qmd
appendix

temp = 0.25
p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
display(Plots.plot(p1, p2, size=(800,320)))
```

>>>>

temp = 0.3
p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
savefig(joinpath(output_images_path, "poc_set_size.png"))
```

### 2023-05-17 notebooks/tables.Rmd
appendix

```{r}
tab_full <- dcast(tab, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = F, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all.tex")
```

@@

```{r}
tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = F, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all_valid.tex")
```

>>>>

```{r}
tab_full <- dcast(tab, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all.tex")
```

@@

```{r}
tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all_valid.tex")
```

### 2023-05-17 paper/paper.tex
appendix

To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}. These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g. crossentropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.

@@

Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance. This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.

@@

In addition to the smooth set size penalty,~\citet{stutz2022learning} also propose a configurable classification loss function, that can be used to enforce coverage. For \textit{MNIST} data, we found that using this function generally improved the visual quality of the generated counterfactuals, so we used it in our experiments involving real-world data. For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples. For this reason we instead relied on standard crossentropy loss for our synthetic datasets. As we have noted in the body of the paper, more experimental work is certainly needed in this context.

>>>>

To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}. These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g. cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.

@@

Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance. This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings. Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
\end{figure}

@@

In addition to the smooth set size penalty,~\citet{stutz2022learning} also propose a configurable classification loss function, that can be used to enforce coverage. For \textit{MNIST} data, we found that using this function generally improved the visual quality of the generated counterfactuals, so we used it in our experiments involving real-world data. For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples. For this reason, we instead relied on standard cross-entropy loss for our synthetic datasets. As we have noted in the body of the paper, more experimental work is certainly needed in this context. Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.

\begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_set_size.png}
  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
\end{figure}
